---
title: "Probability Model Selection using R"
author: "Christopher Martinez Servin"
date: "8/9/2018"
output: github_document
---

This is a simplified guide for selecting and fitting a probability model to a dataset. For detailed information, a good start point is the [Wiki page](https://en.wikipedia.org/wiki/Probability_theory) for *probability theory*.

### Generic Steps
<https://cran.r-project.org/>

1) Compute the moments (mean, std. dev., etc.) from data  
2) Propose a model based on judgment, past experience, and physical constraints  
3) Calibrate the proposed model and fit distribution parameters using techniques such as **Method of Moments**, **Method of Maximum Likelihood**, **Graphical technique 'Probability Plot'**, etc.
4) Make *Qualitatively and Quantitative* comparisons  
       a) Normalized Sample Histogram vs. Fitted PDF
       b) Sample Empirical CDF vs. Fitted CDF
       c) Statistical goodness-of-fit tests such as *$R^2$ for regression*, *$\chi^2$-test*, *K-S test*, etc.

### Exercise One
Let's start by creating our own dataset and using the **fitdistr** function from the **MASS** package to fit the distribution

First we create 250 observations of sample data from a Weibull Distribution  
```{r}
library(MASS)
set.seed(100)
weibull_data <- rweibull(250, shape = 2, scale = 5)
```

Next we visualize the data
```{r}
hist(weibull_data, breaks = 25, main = "Histogram of Weibull Data")
```
Per the generic steps described above, we can propose a model based on judgment. In this case we know the data can fit a weibull distribution
```{r}
fit_weibull = fitdistr(weibull_data, densfun = "weibull")
fit_weibull
```

We know that the parameters used to generate the random data were shape = 2, scale = 5. From the results above, the **fitdistr** function is telling us that the shape is 2.07 and the scale is 4.9. Close enough. Note that **fitdistr** is using the *Method of Maximum Likelihood* to estimate the distribution parameters. To learn more, visit the documentation website -> <https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/fitdistr.html>

We can then plot the histogram for the sample data to the fitted probability density function
```{r}
hist(weibull_data, breaks = 25, probability = TRUE, main = "Histogram vs Fitted Curve Weibull Data")
curve(dweibull(x, fit_weibull$estimate[1], fit_weibull$estimate[2]), col="red", lwd=2, add = T)
```

### Exercise Two
Great, that was a good example. But what if we have no idea of what distribution fits the data? Let's explore the use of a different library: **fitdistrplus**
```{r}
library(fitdistrplus)
```
**fitdistrplus** is another method in R for fitting distributions. For this exercise, lets use the data set *Murders* from the package *dslabs*. This package is put together by [*Rafael Irizarry*](http://rafalab.github.io) from *Harvard University* for the edX online [*HarvardX Data Science program*](https://www.edx.org/professional-certificate/harvardx-data-science)
```{r}
library(dslabs)
data("murders")
```
This data set contains murder data in the United States of America.  

Let's do some data exploration
```{r}
str(murders)
```
```{r}
head(murders)
```

We will work with the total number of murders per U.S. state and try to fit a distribution for this data. Let's visualize it.

```{r}
hist(murders$total,breaks = 20, main = "Histogram of total murders in the U.S. per State")
```
Instead of creating a histogram, we could also plot the empirical probability density and cumulative distribution.
```{r}
plotdist(murders$total, histo = TRUE, demp = TRUE)
```
Another tool is to show descriptive statistics, like the moments, to help us in making a decision
```{r}
descdist(murders$total, discrete = FALSE, boot = 100)
```

Now, before we continue with the interpretation of the results let's make something clear: Real data doesn't necessarily follow any particular distribution. **Parametric distributions are not a perfect description of reality.**

With that being said, the above results are telling us that the skewness and kurtosis are consistent with gamma, exponential, and lognorm distributions. The plot also includes a nonparametric bootstrap procedure for the values of kurtosis and skewness.

#### Fitting a distribution
From the plot above, we choose lognorm, exponential and gamma distributions to fit the data using **fitdist** from **fitdistrplus**.


```{r}
exponential_fit = fitdist(murders$total, "exp")
gamma_fit = fitdist(murders$total, "gamma")
lognormal_fit = fitdist(murders$total, "lnorm")
```
And then we can plot the results
```{r}
plot.legend <- c("Exponential", "Gamma", "Lognormal")
denscomp(list(exponential_fit, gamma_fit, lognormal_fit), legendtext = plot.legend)
cdfcomp (list(exponential_fit, gamma_fit, lognormal_fit), legendtext = plot.legend)
qqcomp  (list(exponential_fit, gamma_fit, lognormal_fit), legendtext = plot.legend)
ppcomp  (list(exponential_fit, gamma_fit, lognormal_fit), legendtext = plot.legend)
```
We know that trying to obtain a solution from a visualization can be subjective, so we use some goodness-of-fit quantitative methods.
```{r}
gofstat(list(exponential_fit, gamma_fit, lognormal_fit), fitnames = c("Exponential", "Gamma", "LogNorm"))
```
From all the above statistics, including K-S test, AIC, BIC, etc, we see that the Gamma function is the best fit for the data. Let's get the parameters of the distribution and plot them against the data.
```{r}
gamma_fit
```
```{r}
hist(murders$total, breaks = 20, probability = TRUE, main = "Histogram vs Fitted Curve Data")
curve(dgamma(x, shape = gamma_fit$estimate[1], rate = gamma_fit$estimate[2]), col="red", lwd=2, add = T)
```
Great! we were able to fit a distribution and obtain the paremeters in order to solve any further questions.
